---
title: "Nutrient Indicators"
execute: 
  warning: false
  error: false
filters: 
  - lightbox
lightbox: auto
bibliography: references.bib
---

```{r}
#| label: data-load
#| include: false
library(here)

source(here::here('R', '00_loadpackages.R'))
source(here::here('R', '02.1_load_wrangle_NUT.R'))
#|source(here::here('R', '02.2_load_wrangle_WQ-and-MET.R'))
source(here::here('R', '03_graph_fxns.R'))

load(here('output','data', 'pi_wq.Rdata'))
load(here('output','data', 'ss_wq.Rdata'))
load(here('output','data', 'fm_wq.Rdata'))
load(here('output','data', 'pc_wq.Rdata'))
```

The state thresholds used in the chlorophyll *a*, total nitrogen, and total phosphorus visualizations come from Florida State Surface Water Quality Standards within the Florida Administrative Code.

::: callout-note
# Access the Standards

By visiting [Florida Rules](https://www.flrules.org/gateway/ruleno.asp?id=62-302.530)
:::

## Summary

Daily average dissolved oxygen values were mostly within or above their average ranges so far in 2024 (@fig-do); however, Pine Island and Pellicer Creek (@fig-do; @fig-do-4) both had above average and below daily average ranges. In previous years, dissolved oxygen values at Pine Island (@fig-do-pi-1) and Pellicer Creek (@fig-do-pc-1) frequently fell below the 2 mg/L hypoxia threshold, but so far in 2024, there have not been any dissolved oxygen readings below that threshold (@fig-do-pi-2; @fig-do-pc-2). 

Chlorophyll *a* values were often higher than average at all stations this year, especially at Pine Island, San Sebastian, and Pellicer Creek (@fig-chla-boxplots). Annual geometric mean (AGM) chlorophyll *a* values were high and above the state threshold levels at all stations, except for Fort Matanzas (@fig-chla-boxplots; @fig-chla-agm; @fig-chla-agm-3). The AGM will likely change for all stations as more data are added throughout the year.

Ammonium (NH4) concentrations for 2024 were below monthly means most of year at all four stations (@fig-nh4-boxplots). Concentrations were above average at the two southern stations, Fort Matanzas and Pellicer Creek in September (@fig-nh4-boxplots-3; @fig-nh4-boxplots-4) but were below average at Pine Island and San Sebastian (@fig-nh4-boxplots-1; @fig-nh4-boxplots-2).

Total Kjeldahl Nitrogen (TKN) concentrations were higher than average at all stations in January and March, but concentrations were lower most of Q2 and Q3 (@fig-tkn-boxplots). Concentrations in September were above average for Fort Matanzas and Pellicer Creek (@fig-tkn-boxplots-3; @fig-tkn-boxplots-4)

Total phosphorus (TP) concentrations were below the state threshold at all stations in 2024 with the exception of February at Pine Island which was missing data (@fig-tp-boxplots). Fort Matanzas experienced higher than average TP values for January -- March and July (@fig-tp-boxplots-3) and San Sebastian had lower than average TP values for February -- September. 

Relatively high levels of orthophosphate (PO4) were observed at Pine Island and San Sebastian in January (@fig-tp-boxplots-1; @fig-tp-boxplots-2) and below average levels were found in April -- June for San Sebastian, Fort Matanzas, and Pellicer Creek (@fig-tp-boxplots-2; (@fig-tp-boxplots-3).

## Data

### Dissolved Oxygen

```{r}
#| label: fig-do
#| echo: false
#| fig.cap: Historical daily range of dissolved oxygen at each SWMP water quality station with current year daily average overlaid in blue. Horizontal dashed red line indicates 2mg/L hypoxia threshold.
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

pi |> 
  filter(do_mgl > 0) |> 
historical_daily_range(param = 'do_mgl', target_yr = 2024, criteria = 2)

historical_daily_range(ss, param = 'do_mgl', target_yr = 2024, criteria = 2)

historical_daily_range(filter(fm, do_mgl > 0), param = 'do_mgl', target_yr = 2024, criteria = 2)

pc |>
  filter(do_mgl < 20 & do_mgl > 0) |> 
  historical_daily_range(param = 'do_mgl', target_yr = 2024, criteria = 2)

```

Dissolved oxygen at Pine Island has been uncharacteristically low in the past two years.

```{r}
#| label: fig-do-pi
#| echo: false
#| fig-cap: Continuous dissolved oxygen at Pine Island with monthly average overlaid in solid gray line. Conditions are indicated as good when above 5 mg/L, fair between 2 and 5 mg/L, and poor if less than 2 mg/L.
#| fig.subcap:
#|     - "For the past five years"
#|     - "For the current 2024 year"
#| layout-ncol: 2

threshold_criteria_plot(pi, param = 'do_mgl', rng = c(2019, 2024), 
                        thresholds = c(2,5),
                        threshold_labs = c('Poor', 'Fair', 'Good'), 
                        monthly_smooth = TRUE, 
                        threshold_cols = c('#FEC596', '#FFFFCC', '#ABD9E9')) +
  theme(axis.text = element_text(color = "black", size = 12),
        axis.title.y = element_text(size = 14))

threshold_criteria_plot(pi, param = 'do_mgl', rng = c(2024, 2024), 
                        thresholds = c(2,5),
                        threshold_labs = c('Poor', 'Fair', 'Good'), 
                        monthly_smooth = TRUE, 
                        threshold_cols = c('#FEC596', '#FFFFCC', '#ABD9E9')) +
  theme(axis.text = element_text(color = "black", size = 12),
        axis.title.y = element_text(size = 14))

```

Pellicer Creek is located within a waterbody with known impairments for dissolved oxygen, chlorophyll *a*, and fecal coliform bacteria. As such, this station frequently experiences low dissolved oxygen values, particularly in the summer months (@fig-do-pc).

```{r}
#| label: fig-do-pc
#| echo: false
#| fig.cap: Continuous dissolved oxygen at Pellicer Creek with monthly average overlaid in solid gray line. Conditions are indicated as good when above 5 mg/L, fair between 2 and 5 mg/L, and poor if less than 2 mg/L.
#| fig.subcap:
#|     - "For the past five years"
#|     - "For the current 2024 year"
#| layout-ncol: 2
    
threshold_criteria_plot(pc, param = 'do_mgl', rng = c(2019, 2024), 
                        thresholds = c(2,5),
                        threshold_labs = c('Poor', 'Fair', 'Good'), 
                        monthly_smooth = TRUE, 
                        threshold_cols = c('#FEC596', '#FFFFCC', '#ABD9E9')) +
  theme(axis.text = element_text(color = "black", size = 12),
        axis.title.y = element_text(size = 14))



threshold_criteria_plot(pc, param = 'do_mgl', rng = c(2024, 2024), 
                        thresholds = c(2,5),
                        threshold_labs = c('Poor', 'Fair', 'Good'), 
                        monthly_smooth = TRUE, 
                        threshold_cols = c('#FEC596', '#FFFFCC', '#ABD9E9')) +
  theme(axis.text = element_text(color = "black", size = 12),
        axis.title.y = element_text(size = 14))

```



### Chlorophyll *a*

```{r}
#| label: fig-chla-boxplots
#| echo: false
#| fig.cap: Historical monthly chlorophyll *a* at each SWMP water quality station with current year monthly value overlaid in red. Horizontal dashed blue line indicates the state threshold, as found in F.A.C 62-302.530.
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

boxplot_currentyear(station = "gtmpinut", 
                     param = chla_n, 
                     threshold = TRUE,
                     thres_val = 6.6) +
  labs(y = "Mean Monthly Chlorophyll-a (\U00B5g/L)") +
  annotate("text",
             x = "Jun",
             y = 20,
             size = 3,
             color = "blue",
             label = "State Threshold 6.6 (\U00B5g/L)")

boxplot_currentyear(station = "gtmssnut", 
                     param = chla_n, 
                     threshold = TRUE,
                     thres_val = 4.0) +
  labs(y = "Mean Monthly Chlorophyll-a (\U00B5g/L)") +
  annotate("text",
           x = "Jun",
           y = 16,
           size = 3,
           color = "blue",
           label = "State Threshold 4.0 (\U00B5g/L)")

boxplot_currentyear(station = "gtmfmnut", 
                    param = chla_n, 
                    threshold = TRUE,
                    thres_val = 5.5) +
  labs(y = "Mean Monthly Chlorophyll-a (\U00B5g/L)") +
  annotate("text",
           x = "Jun",
           y = 24,
           size = 3,
           color = "blue",
           label = "State Threshold 5.5 (\U00B5g/L)")

boxplot_currentyear(station = "gtmpcnut", 
                     param = chla_n, 
                     threshold = TRUE,
                     thres_val = 4.3) +
  labs(y = "Mean Monthly Chlorophyll-a (\U00B5g/L)") +
  annotate("text",
           x = "Jun",
           y = 38,
           size = 3,
           color = "blue",
           label = "State Threshold 4.3 (\U00B5g/L)")
```

```{r}
#| label: fig-chla-agm
#| echo: false
#| fig.cap: Annual geometric means of monthly chlorophyll *a* at each SWMP water quality station. Horizontal dashed blue line indicates the state threshold, as found in F.A.C 62-302.530. Values exceeding the threshold are indicated in red. The 2024 value is based upon data to date and does not equal an entire year of data.
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

agm(station = "gtmpinut", 
    param = CHLA_agm, 
    threshold = TRUE,
    thres_val = 6.6) +
  labs(y = "Geo.Mean Annual Chlorophyll-a (\U00B5g/L)") +
  annotate("text",
           x = "2006",
           y = 8,
           size = 3,
           color = "blue",
           label = "State Threshold 6.6 (\U00B5g/L)")

agm(station = "gtmssnut", 
    param = CHLA_agm, 
    threshold = TRUE,
    thres_val = 4.0) +
  labs(y = "Geo.Mean Annual Chlorophyll-a (\U00B5g/L)") +
  annotate("text",
           x = "2006",
           y = 6,
           size = 3,
           color = "blue",
           label = "State Threshold 4.0 (\U00B5g/L)")

agm(station = "gtmfmnut", 
    param = CHLA_agm, 
    threshold = TRUE,
    thres_val = 5.5) +
  labs(y = "Geo.Mean Annual Chlorophyll-a (\U00B5g/L)") +
  annotate("text",
           x = "2006",
           y = 6,
           size = 3,
           color = "blue",
           label = "State Threshold 5.5 (\U00B5g/L)")

agm(station = "gtmpcnut", 
    param = CHLA_agm, 
    threshold = TRUE,
    thres_val = 4.3) +
  labs(y = "Geo.Mean Annual Chlorophyll-a (\U00B5g/L)") +
  annotate("text",
           x = "2006",
           y = 16,
           size = 3,
           color = "blue",
           label = "State Threshold 4.3 (\U00B5g/L)")

```

### Nitrogen

#### Total Nitrogen

Total Nitrogen (TN) is calculated by adding TKN and NO23. It accounts for the inorganic nitrogen fraction as well as larger pools of organic bioavailable forms.

Several of the NO23 values fell below the Florida Department of Environmental Protection Central Lab's minimum detection limit and therefore TN was not able to be calculated and is missing for quite a few months at all the stations (@fig-tn-boxplots). This was especially true at Fort Matanzas which has the least amount of TN data (@fig-tn-boxplots-3).

```{r}
#| label: fig-tn-boxplots
#| echo: false
#| fig.cap: Historical monthly total nitrogen at each SWMP water quality station with current year monthly value overlaid in red. Horizontal dashed blue line indicates the state threshold, as found in F.A.C 62-302.530.
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

boxplot_currentyear(station = "gtmpinut",
                     param = tn,
                     threshold = TRUE,
                     thres_val = 0.65) + 
  labs(y = "Mean Monthly Total Nitrogen (mg/L)") +
  annotate("text", 
           x = "Jun",
           y = 1.2,
           size = 3,
            color = "blue",
            label = "State Threshold 0.65 (mg/L)")

 boxplot_currentyear(station = "gtmssnut",
                     param = tn,
                     threshold = TRUE,
                     thres_val = 0.55) +
   labs(y = "Mean Monthly Total Nitrogen (mg/L)") +
   annotate("text",
            x = "Jun", 
            y = 1.0,
            size = 3,
            color = "blue",
            label = "State Threshold 0.55 (mg/L)")

 boxplot_currentyear(station = "gtmfmnut",
                      param = tn,
                      threshold = TRUE,
                      thres_val = 0.55) +
   labs(y = "Mean Monthly Total Nitrogen (mg/L)") +
   annotate("text",
            x = "Jun",
            y = 1.0,
            size = 3,
            color = "blue",
            label = "State Threshold 0.55 (mg/L)")

 boxplot_currentyear(station = "gtmpcnut",
                      param = tn,
                      threshold = TRUE,
                      thres_val = 1.10) +
   labs(y = "Mean Monthly Total Nitrogen (mg/L)") +
   annotate("text",
            x = "Jun",
            y = 1.7,
            size = 3,
            color = "blue",
            label = "State Threshold 1.10 (mg/L)")
```

#### Ammonium

Ammonium (NH4) is directly measured in our samples and is a key inorganic nutrient assimilated by phytoplankton in primary production. 

```{r}
#| label: fig-nh4-boxplots
#| echo: false
#| fig.cap: Historical monthly ammonium at each SWMP water quality station with current year monthly value overlaid in red. 
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

boxplot_currentyear(station = "gtmpinut", 
                     param = nh4f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Ammonium (mg/L)") 

boxplot_currentyear(station = "gtmssnut", 
                     param = nh4f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Ammonium (mg/L)")

boxplot_currentyear(station = "gtmfmnut", 
                     param = nh4f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Ammonium (mg/L)")

boxplot_currentyear(station = "gtmpcnut", 
                     param = nh4f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Ammonium (mg/L)")


```

#### Nitrite + Nitrate

```{r}
#| label: fig-no23-boxplots
#| echo: false
#| fig.cap: Historical monthly nitrite + nitrate at each SWMP water quality station with current year monthly value overlaid in red. 
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

boxplot_currentyear(station = "gtmpinut", 
                     param = no23f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Nitrite + Nitrate (mg/L)") 

boxplot_currentyear(station = "gtmssnut", 
                     param = no23f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Nitrite + Nitrate (mg/L)")

boxplot_currentyear(station = "gtmfmnut", 
                     param = no23f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Nitrite + Nitrate (mg/L)")

boxplot_currentyear(station = "gtmpcnut", 
                     param = no23f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Nitrite + Nitrate (mg/L)")


```

#### Total Kjeldahl Nitrogen

TKN is the total concentration of organic nitrogen and ammonia. It is directly measured in our samples.

```{r}
#| label: fig-tkn-boxplots
#| echo: false
#| fig.cap: Historical monthly total kjeldahl nitrogen at each SWMP water quality station with current year monthly value overlaid in red. 
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

boxplot_currentyear(station = "gtmpinut", 
                     param = tkn, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Total Kjeldahl Nitrogen (mg/L)") 

boxplot_currentyear(station = "gtmssnut", 
                     param = tkn, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Total Kjeldahl Nitrogen (mg/L)")

boxplot_currentyear(station = "gtmfmnut", 
                     param = tkn, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Total Kjeldahl Nitrogen (mg/L)")

boxplot_currentyear(station = "gtmpcnut", 
                     param = tkn, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Total Kjeldahl Nitrogen (mg/L)")


```

### Phosphorus

#### Total Phosphorus

```{r}
#| label: fig-tp-boxplots
#| echo: false
#| fig.cap: Historical monthly total phosphorus at each SWMP water quality station with current year monthly value overlaid in red. Horizontal dashed blue line indicates the state threshold, as found in F.A.C 62-302.530.
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

boxplot_currentyear(station = "gtmpinut", 
                    param = tp, 
                    threshold = TRUE,
                    thres_val = 0.105) +
  labs(y = "Mean Monthly Total Phosphorus (mg/L)") +
  annotate("text",
           x = "Mar",
           y = 0.15,
           size = 3,
           color = "blue",
           label = "State Threshold 0.105 (mg/L)")

boxplot_currentyear(station = "gtmssnut",
                    param = tp,
                    threshold = TRUE, 
                    thres_val = 0.11) +
  labs(y = "Mean Monthly Total Phosphorus (mg/L)") +
  annotate("text",
           x = "Mar",
           y = 0.12,
           size = 3,
           color = "blue",
           label = "State Threshold 0.11 (mg/L)")

boxplot_currentyear(station = "gtmfmnut", 
                     param = tp,
                     threshold = TRUE, 
                     thres_val = 0.111) +
  labs(y = "Mean Monthly Total Phosphorus (mg/L)") +
  annotate("text",
           x = "Jun",
           y = 0.12,
           size = 3,
           color = "blue",
           label = "State Threshold 0.111 (mg/L)")

boxplot_currentyear(station = "gtmpcnut", 
                     param = tp, 
                     threshold = TRUE,
                     thres_val = 0.123) +
  labs(y = "Mean Monthly Total Phosphorus (mg/L)") +
  annotate("text",
           x = "Jun",
           y = 0.25,
           size = 3,
           color = "blue",
           label = "State Threshold 0.123 (mg/L)")
```

#### Orthophosphate

```{r}
#| label: fig-po4-boxplots
#| echo: false
#| fig.cap: Historical monthly orthophosphate at each SWMP water quality station with current year monthly value overlaid in red.
#| fig.subcap:
#|    - "Pine Island"
#|    - "San Sebastian"
#|    - "Fort Matanzas"
#|    - "Pellicer Creek"
#| layout-ncol: 2

boxplot_currentyear(station = "gtmpinut", 
                     param = po4f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Orthophosphate (mg/L)") 

boxplot_currentyear(station = "gtmssnut", 
                     param = po4f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Orthophosphate (mg/L)") 

boxplot_currentyear(station = "gtmfmnut", 
                     param = po4f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Orthophosphate (mg/L)") 

boxplot_currentyear(station = "gtmpcnut", 
                     param = po4f, 
                     threshold = FALSE) +
 labs(y = "Mean Monthly Orthophosphate (mg/L)") 
```
